---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.dim=c(4, 2.5), fig.align='center')
```

```{r, echo=FALSE}
library(ggplot2)
library(bayestestR)
library(knitr)
```


# 2.Log-normal distribution and the Gini coefficient 

We consider a sample of 8 observations representing monthly incomes (in thousands of Swedish Krona). Assume the data follow a log-normal distribution:

$$y_1, ..., y_n|\mu, \sigma^2 \sim \log N(\mu, \sigma^2)$$

with known $\mu = 3.65$ and unknown variance $\sigma^2$. For the variance, we use a non-informative prior:

$$p(\sigma^2) \propto 1/\sigma^2$$

Under this model, the posterior distribution for $\sigma^2$ follows a scaled inverse-chi-squared distribution:

$$\text{Scale-inv-}\chi^2(n, \tau^2), \quad  \text{where} \quad \tau = \frac{\sum^n_{i = 1}(\log y_i - \mu)^2}{n}$$

To draw samples from this posterior, we use the fact that if:

$$\sigma \sim \text{Scale-inv-}\chi^2(n, \tau^2), \quad \text{then} \quad \sigma^2 = \frac{n\tau^2}{\chi^2_n}$$

We generate 10000 posterior draws for $\sigma^2$ and visualize the resulting distribution below.

```{r, echo=FALSE}
y <- c(22, 33, 31, 29, 65, 78, 17, 24)
n <- 8
mu <- 3.65
tau <- sum((log(y) - mu)^2)/n

set.seed(12345)
samples <- (n * tau) / rchisq(10000, df = n)
ggplot(mapping = aes(x = samples)) +
  geom_histogram(bins = 50, fill = "#4E79A7", color = "white", alpha = 0.8) +
  labs(
    title = expression(paste("Posterior of ", sigma^2)),
    x = expression(sigma^2),
    y = "Frequency"
  ) +
  theme_minimal()
```

Then, the Gini coefficient is calculated using the formula:
$$G = 2\Phi(\sigma/\sqrt2) - 1$$
where $\Phi$ denotes the cumulative distribution function of the standard normal distribution, and $\sigma = \sqrt{\sigma^2}$. Using the posterior draws of $\sigma^2$, we obtain the posterior distribution of the Gini coefficient:

```{r, echo=FALSE}
gini <- 2 * pnorm(sqrt(samples)/sqrt(2), 0, 1) - 1
ggplot(mapping = aes(x = gini)) +
  geom_histogram(bins = 50, fill = "#4E79A7", color = "white", alpha = 0.8) +
  labs(
    title = "Posterior of G",
    x = expression(G),
    y = "Frequency"
  ) +
  theme_minimal()
```

Using the posterior distribution of the Gini coefficient $G$, we compute both a 95% equal-tail credible interval and a 95%  Highest Posterior Density Interval (HPDI) for $G$.

```{r, echo=FALSE}
ci <- quantile(gini, c(0.025, 0.975))
hpdi <- hdi(gini, ci = 0.95)
table <- data.frame("Lower bound" = c(ci[1], hpdi[1,2]), "Upper bound" = c(ci[2], hpdi[1,3]))
rownames(table) <- c("Equal tail", "HPDI")
kable(table, caption = "Confidence intervals for G")
```

The equal-tail interval removes 2.5% of the posterior mass from each tail, while the HPDI represents the narrowest interval containing 95% of the posterior mass. In this case, the HPDI is slightly narrower and starts at a lower bound compared to the equal-tail interval. This suggests that the posterior distribution of $G$ is slightly right-skewed, and the HPDI better captures the region of highest probability. Therefore, if interpretability and precision are important, the HPDI may be preferred.